---
title: "Session 4 lab exercise: Poisson log-linear regression"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
clean: false
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
  md_document:
    preserve_yaml: false
always_allow_html: true
---

**Learning objectives**

1. Simulate Poisson-distributed data with a relevant covariate
2. Fit a Poisson log-linear GLM
3. Create and interpret diagnostic plots for a log-linear GLM
4. Use analysis of deviance to compare two log-linear GLMs
5. Practice recoding and creating tables and plots


**Activities**

## 1. Simulate count data from a Poisson distribution

(For example, number of hospital visits by persons over 70 in a 3-year period), where:
    a. 10,000 persons annotated with "race" as "white" or "non-white"
    b. "white" persons have an average of 3.5 hospital visits during this time period
    c. "non-white" persons have an average of 3 hospital visits

```{r}
library(tidyverse)
```

```{r}
set.seed(1)
N <- 10000
simdat <- data.frame(race = sample(c("white", "non-white"), N, replace = TRUE)) %>%
  mutate(race = factor(race, levels = c("white", "non-white"))) %>%
  mutate(y = rpois(N, lambda = ifelse(race == "white", 3.5, 3.0)))
```




## 2. Fit a log-linear Poisson model of count outcomes

With "race" as the predictor. Note, in this context I tend to use the terms "predictor" and "covariate" interchangeably, to mean any variable used as a predictor in the regression model.

```{r}
fit <- glm(y ~ race, data = simdat, family = poisson(link = "log"))
summary(fit)
```

## 3. Use a chi-square test on deviance residuals

To test null hypothesis of no relationship between mean hospital visits and race.


* The difference in total deviance between two nested models is $\chi^2$ distributed under $H_0$ that the more complex model is no better at explaining the response.
     + The difference in deviance residuals is (11041 - 10831) = 210, with a difference of 1 degrees of freedom.

The critical threshold for rejection at p=0.05 is:
```{r}
qchisq(0.95, df = 1)
```

So we reject $H_0$

BEWARE OF MISSING DATA: THIS IS SAFER

```{r}
fit0 <- glm(y ~ 1, data = simdat, family = poisson(link = "log"))
anova(fit0, fit, test = "LRT")
```

## 4. Create and discuss standard fit diagnostics plots

```{r, echo=FALSE}
par(mfrow = c(2, 2))
plot(fit)
```

## 5. Example: Risky Drug Use Behavior

* Download the "needle_sharing" dataset (see Vittinghoff 8.3.1)
* Outcome is # times the drug user shared a syringe in the past month (`shared_syr`)
* Predictors: sex, ethn, homeless

```{r loaddata}
library(readxl)
needledat <- read_excel("needle_sharing.xlsx")
summary(needledat$shared_syr)
var(needledat$shared_syr, na.rm = TRUE)
```

Some recoding:
```{r recode}
suppressPackageStartupMessages(library(dplyr))
needledat_cleaned <-
  mutate(needledat,
    homeless = factor(homeless, levels = 0:1, labels = c("No", "Yes")),
    sex = factor(sex, levels = c("M", "F"), labels = c("Male", "Female")),
    ethnicity = factor(ethn)
  ) %>%
  select(all_of(c("shared_syr", "ethnicity", "sex", "homeless")))
```

## 6. Create a table of the risky drug use behavior dataset

```{r table1}
library(table1)
table1(~., data = needledat_cleaned)
```

## 7. Plots of Risky Drug Use Behavior

a. Create a histogram number of syringe uses

b. Create a scatter plot of number of syringe uses versus rank of number of syringe uses

```{r histogram}
library(ggplot2)
ggplot(needledat, aes(shared_syr)) +
  geom_histogram() +
  labs(title = "Counts of Syringe Sharing Incidents Per Person") +
  xlab("Number of Incidents") +
  ylab("Number of people with that count")
```

```{r rankplot}
library(dplyr)
mutate(needledat, rnk = rank(shared_syr, ties.method = "first")) %>%
  ggplot(aes(x = rnk, y = shared_syr)) +
  geom_point() +
  labs(title = "Count vs Rank Count of Syringe Sharing Incidents") +
  xlab("rank of count") +
  ylab("count of syringe sharing")
```

* There are a _lot_ of zeros - Poisson model is not a good fit

## 8. Fit a Poisson model to the risky drug use behavior dataset anyways

Even though we know it is a bad fit

```{r glmfit}
fit.pois <- glm(shared_syr ~ sex + ethn + homeless,
  data = needledat,
  family = poisson(link = "log")
)
summary(fit.pois)
```

## 9. Create and discuss Poisson model diagnostic plots

```{r diagnostic}
par(mfrow = c(2, 2))
plot(fit.pois)
```

## Interpreting Poisson Coefficients

In a Poisson log-linear model, exponentiating the coefficients ($\exp(\beta)$) yields the Incidence Rate Ratios (IRRs). For example, if we exponentiate the coefficients from `fit.pois`:

```{r}
# Exponentiate the coefficients to get IRRs
exp(coef(fit.pois))
```
This tells us the multiplicative effect of each predictor on the expected count of syringe sharing incidents, holding other variables constant.

# Exercises

## Exercise 1: Calculate Incidence Rate Ratios (IRRs)
Extract the coefficients from `fit.pois` (the model on risky drug use), exponentiate them to obtain the Incidence Rate Ratios (IRRs), and compute their 95% confidence intervals.

```{r}

```

## Exercise 2: Model Comparison (LRT)
Fit a reduced Poisson model by removing the `homeless` predictor from the previous risky drug use model. Then, perform a Likelihood Ratio Test (using `anova(..., test="LRT")`) to compare this reduced model against the full model `fit.pois`. Interpret the result. Does the `homeless` variable significantly improve the model fit?

```{r}

```

## Exercise 3: Check for Overdispersion
Calculate the ratio of the residual deviance to the residual degrees of freedom for `fit.pois`. Interpret what this value means for the model assumptions. Is there evidence of overdispersion?

```{r}

```
