---
title: "Lecture and lab materials for CUNY SPH Biostatistics 2"
author: "Levi Waldron"
output: 
  md_document:
    preserve_yaml: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start -->
[![pages-build-deployment](https://github.com/waldronbios2/cunybios2/actions/workflows/pages/pages-build-deployment/badge.svg)](https://github.com/waldronbios2/cunybios2/actions/workflows/pages/pages-build-deployment)
[![Website and Docker](https://github.com/waldronbios2/cunybios2/actions/workflows/build.yaml/badge.svg)](https://github.com/waldronbios2/cunybios2/actions/workflows/build.yaml)
<!-- badges: end -->

# Overview

This course is designed to expand on the biostatistical methods
covered in Applied Biostatistics II. It introduces generalized linear
regression including logistic regression, loglinear models (including
Poisson and Negative Binomial models of count data), survival
analysis, and longitudinal data including random and mixed effect
models.

# Materials

```{r youtube, echo=FALSE}
chapters <- dir(pattern = "session*")
chapters <- chapters[order(as.integer(sub("session", "", chapters)))]
youtube <- read.csv("youtubeURLs.csv")[1:length(chapters), ]
youtube$lecturetips <-
  ifelse(youtube$lecture == "#", "not yet available", "watch on YouTube")
youtube$labtips <-
  ifelse(youtube$lab == "#", "not yet available", "watch on YouTube")
```

```{r packages, echo=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(kableExtra)
})
```

```{r, echo=FALSE}
dt <- data.frame(
  session = cell_spec(
    seq_along(chapters),
    tooltip = sapply(chapters, function(x) {
      read.dcf(file.path(x, "DESCRIPTION"))[1, "Description"]
    }, USE.NAMES = FALSE),
    link = paste0("https://github.com/waldronbios2/cunybios2/tree/main/", chapters)
  ),
  topic = sapply(chapters, function(x) {
    read.dcf(file.path(x, "DESCRIPTION"))[1, "Description"]
  }, USE.NAMES = FALSE),
  git = cell_spec(
    "source",
    tooltip = "browse session directory in monorepo",
    link = paste0("https://github.com/waldronbios2/cunybios2/tree/main/", chapters)
  ),
  lecture = cell_spec(
    "HTML",
    tooltip = "lecture notes in HTML",
    link = paste0(
      "https://waldronbios2.github.io/cunybios2/",
      chapters,
      "/articles/session_lecture.html"
    )
  ),
  lecturepdf = cell_spec(
    "PDF",
    tooltip = "lecture notes in PDF",
    link = paste0(
      "https://waldronbios2.github.io/cunybios2/",
      chapters,
      "/articles/session_lecture.pdf"
    )
  ),
  lecturevideo = cell_spec(
    "watch",
    tooltip = youtube$lecturetips,
    link = youtube$lecture[1:length(chapters)]
  ),
  labhtml = cell_spec(
    "HTML",
    tooltip = "lab notes in HTML",
    link = paste0(
      "https://waldronbios2.github.io/cunybios2/",
      chapters,
      "/articles/session_lab.html"
    )
  ),
  labvideo = cell_spec("watch",
                       tooltip = youtube$labtips,
                       link = youtube$lab[1:length(chapters)]),
  row.names = NULL
)
```

```{r maketable, echo=FALSE}
kableExtra::kbl(
  dt,
  format = "html",
  escape = FALSE,
  col.names = NULL,
  align = "c",
  row.names = FALSE
) %>%
  kable_paper() %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::add_header_above(c(
    "Session" = 1,
    "Topic" = 1,
    "GitHub" = 1,
    "Lecture" = 3,
    "Lab" = 2
  ))
```

# Cloning course materials (for students)

You can download course materials individually from this website, but it is more convenient to clone all course materials at once. However, cloning requires a system installation of `git`, and the installation methods depend on your operating system. Here are [git installation instructions for RStudio](https://support.rstudio.com/hc/en-us/articles/200532077-Version-Control-with-Git-and-SVN). If you are using RStudio on https://anvil.terra.bio (ask the professor for access with free compute), `git` is pre-installed so you don't have to install it. Once you have git working from RStudio, you can do the following to obtain a local copy of all course materials:

1. In RStudio, select "File - New Project".
2. Choose "Version Control" then "Git". 
3. Enter the URL: `https://github.com/waldronbios2/cunybios2.git`. Enter any directory name (e.g. "cunybios2"), and click "Browse" to choose the current directory. Then click "Create Project".

That's it! You can now find all 10 sessions worth of lecture and lab code in the project files. Open any session folder (e.g. `session1/vignettes/session_lecture.Rmd`) and click "Knit" to build the lecture slides or lab notes.

# Data Science Units

This course is supported by free access to
[DataCamp](https://www.datacamp.com/) for registered students.

# Technical notes for teachers

To add a new session to this monorepo:

1. Create a new directory `sessionXX` with the session package contents (`DESCRIPTION`, `vignettes/`, etc.).
2. Add the new session to the `Suggests` and `Remotes` sections of the top-level `DESCRIPTION` file.
3. Update `youtubeURLs.csv` with links to lecture and lab recordings.
4. Commit and push:
```bash
git add .
git commit -m "add sessionXX materials"
git push origin main
```

