---
title: "Lecture and lab materials for CUNY SPH Biostatistics 2"
author: "Levi Waldron"
output: 
  md_document:
    preserve_yaml: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start --> ![Website and Docker](https://github.com/waldronbios2/cunybios2/workflows/Website%20and%20Docker/badge.svg) <!-- badges: end -->

This course is designed to expand on the biostatistical methods
covered in Applied Biostatistics I. It introduces generalized linear
regression including logistic regression, loglinear models (including
Poisson and Negative Binomial models of count data), survival
analysis, and longitudinal data including random and mixed effect
models.

```{r youtube, echo=FALSE}
chapters <- dir(pattern = "session*")
youtube <- read.csv("youtubeURLs.csv")[1:length(chapters), ]
youtube$lecturetips <-
  ifelse(youtube$lecture == "#", "not yet available", "watch on YouTube")
youtube$labtips <-
  ifelse(youtube$lab == "#", "not yet available", "watch on YouTube")
```

```{r, echo=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(kableExtra)
})
badgeimages <- paste0("https://github.com/waldronbios2/", chapters, 
                 "/workflows/build/badge.svg")
badgeimagehtml <- paste0("<img src='", badgeimages, "'>")
badgelinks <- paste0("https://github.com/waldronbios2/", chapters, "/actions")
badgeall <- paste0("<a href = '", badgelinks, "'>", badgeimagehtml, "</a>")
dt <- data.frame(
  session = cell_spec(
    seq_along(chapters),
    tooltip = sapply(chapters, function(x)
      packageDescription(x)$Description),
    link = paste0("https://waldronbios2.github.io/", chapters)
  ),
  status = badgeall,
  git = cell_spec(
  "git clone",
  tooltip = "use for git clone click to go to GitHub repo",
  link = paste0("https://github.com/waldronbios2/", chapters, ".git")
),
lecture = cell_spec(
  "HTML",
  tooltip = "lecture notes in HTML",
  link = paste0(
    "https://waldronbios2.github.io/",
    chapters,
    "/articles/session_lecture.html"
  )
),
lecturepdf = cell_spec(
  "PDF",
  tooltip = "lecture notes in PDF",
  link = paste0(
    "https://waldronbios2.github.io/",
    chapters,
    "/articles/session_lecture.pdf"
  )
),
lecturevideo = cell_spec(
  "watch",
  tooltip = youtube$lecturetips,
  link = youtube$lecture[1:length(chapters)]
  ),
labhtml = cell_spec(
  "HTML",
  tooltip = "lab notes in HTML",
  link = paste0(
    "https://waldronbios2.github.io/",
    chapters,
    "/articles/session_lab.html"
    )
  ),
labvideo = cell_spec(
  "watch",
  tooltip = youtube$labtips,
  link = youtube$lab[1:length(chapters)]
  )
)

kableExtra::kbl(
  dt,
  format = "html",
  escape = FALSE,
  col.names = NULL,
  align = "c"
) %>%
  kable_paper() %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::add_header_above(c(
    "Session" = 1,
    "Status" = 1,
    "GitHub" = 1,
    "Lecture" = 3,
    "Lab" = 2
  ))
```
