---
title: "Lecture and lab materials for CUNY SPH Biostatistics 2"
author: "Levi Waldron"
output: 
  md_document:
    preserve_yaml: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start --> ![Website and Docker](https://github.com/waldronbios2/cunybios2/workflows/Website%20and%20Docker/badge.svg) <!-- badges: end -->

# Overview

This course is designed to expand on the biostatistical methods
covered in Applied Biostatistics I. It introduces generalized linear
regression including logistic regression, loglinear models (including
Poisson and Negative Binomial models of count data), survival
analysis, and longitudinal data including random and mixed effect
models.

# Materials

```{r youtube, echo=FALSE}
chapters <- dir(pattern = "session*")
chapters <- chapters[order(as.integer(sub("session", "", chapters)))]
youtube <- read.csv("youtubeURLs.csv")[1:length(chapters), ]
youtube$lecturetips <-
  ifelse(youtube$lecture == "#", "not yet available", "watch on YouTube")
youtube$labtips <-
  ifelse(youtube$lab == "#", "not yet available", "watch on YouTube")
```

```{r packages, echo=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(kableExtra)
})
```

```{r links, echo=FALSE}
badgeimages <- paste0("https://github.com/waldronbios2/", chapters, 
                 "/workflows/build/badge.svg")
badgeimagehtml <- paste0("<img src='", badgeimages, "'>")
badgelinks <- paste0("https://github.com/waldronbios2/", chapters, "/actions")
badgeall <- paste0("<a href = '", badgelinks, "'>", badgeimagehtml, "</a>")
```

```{r, echo=FALSE}
dt <- data.frame(
  session = cell_spec(
    seq_along(chapters),
    tooltip = sapply(chapters, function(x) {
      pd <- packageDescription(x, lib.loc=".")
      if (!is(pd, "packageDescription")){
        warning(x, " is not a package")
	return(NA)
      }
      return(packageDescription(x, lib.loc=".")$Description)
    }),
    link = paste0("https://waldronbios2.github.io/", chapters)
  ),
  status = badgeall,
  git = cell_spec(
    "git clone",
    tooltip = "use for git clone click to go to GitHub repo",
    link = paste0("https://github.com/waldronbios2/", chapters, ".git")
  ),
  lecture = cell_spec(
    "HTML",
    tooltip = "lecture notes in HTML",
    link = paste0(
      "https://waldronbios2.github.io/",
      chapters,
      "/articles/session_lecture.html"
    )
  ),
  lecturepdf = cell_spec(
    "PDF",
    tooltip = "lecture notes in PDF",
    link = paste0(
      "https://waldronbios2.github.io/",
      chapters,
      "/articles/session_lecture.pdf"
    )
  ),
  lecturevideo = cell_spec(
    "watch",
    tooltip = youtube$lecturetips,
    link = youtube$lecture[1:length(chapters)]
  ),
  labhtml = cell_spec(
    "HTML",
    tooltip = "lab notes in HTML",
    link = paste0(
      "https://waldronbios2.github.io/",
      chapters,
      "/articles/session_lab.html"
    )
  ),
  labvideo = cell_spec("watch",
                       tooltip = youtube$labtips,
                       link = youtube$lab[1:length(chapters)])
)
```

```{r maketable, echo=FALSE}
kableExtra::kbl(
  dt,
  format = "html",
  escape = FALSE,
  col.names = NULL,
  align = "c"
) %>%
  kable_paper() %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::add_header_above(c(
    "Session" = 1,
    "Status" = 1,
    "GitHub" = 1,
    "Lecture" = 3,
    "Lab" = 2
  ))
```

# Technical notes

To add a new git submodule to this repo:

First, add the new session to the Suggests and Remotes sections of the DESCRIPTION file.

Second, invoke the following from within the local waldronbios2 clone (example for session10):
```
git submodule add https://github.com/waldronbios2/session10.git
git commit -am "add session10 submodule"
```

Finally, update the file `youtubeURLs.csv` with links to lecture and lab recordings, then push changes:
```
git commit -am "add session10 recording links"
git push origin master
```
