---
title: "Sessions 3 lab exercise: Regression analysis of contraceptive use data"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
clean: false
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
  md_document:
    preserve_yaml: false
always_allow_html: true    
---

**Learning objectives**

1. perform and interpret logistic regression
    + interpret logistic regression coefficients
    + make predictions based on a logistic regression model
2. perform and interpret likelihood ratio test

**Exercises**

1. What is the mean fraction of women using birth control for each age group? Each education level? For women who do or don't want more children?
     - Hint: look at the "data wrangling" cheat sheet functions `mutate`, `group_by`, and `summarize`
2. Based on ```fit1```, write on paper the model for expected probability of using birth control.  Don't forget the logit function.
3. Based on ```fit1```, what is the expected probability of an individual 25-29 years old, with high education, who wants more children, using birth control? Calculate it manually, and using `predict(fit1)`
4. Based on ```fit1```: Relative to women under 25 who want to have children, what is the predicted increase in odds that a woman 40-49 years old who does _not_ want to have children will be taking birth control?
5. Using a likelihood ratio test, is there evidence that a model with interactions improves on ```fit1``` (no interactions)?
6. Which, if any, variables have the strongest interactions?
7. Looking at the effect of age only, consider contrasts between *every pair* of age groups. Between which age groups is the contrast significant?

# Load and explore the contraceptive use data

Load the data from http://data.princeton.edu/wws509/datasets/#cuse. From this page:
> These data show the distribution of 1607 currently married and fecund women interviewed in the Fiji Fertility Survey, according to age, education, desire for more children and current use of contraception.

See Session 2 for some discussion about reading this file. Here we just do it.

```{r, message=FALSE}
library(readr)
cuse <- read_table("cuse.dat",
  col_types = cols(
    age = col_factor(levels = c("<25", "25-29", "30-39", "40-49")),
    education = col_factor(levels = c("low", "high")),
    wantsMore = col_factor(levels = c("no", "yes")),
    X6 = col_skip()
  )
)
summary(cuse)
```

# What is the mean fraction of women using birth control for...

What is the mean fraction of women using birth control for each age group? Each education level? For women who do or don't want more children?
     - Hint: look at the ["Data Transformation Cheatsheet"](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf) functions `mutate`, `group_by`, and `summarize`. This is one of several very handy cheatsheets produced
     by RStudio, see https://rstudio.com/resources/cheatsheets/ for a list.
     

**Solution**

First create a new column containing the fraction using contraception:

```{r, message=FALSE}
library(dplyr)
cuse2 <- mutate(cuse, fracusing = using / (using + notUsing))
cuse2
```

Now, group and summarize by age:

```{r}
cuse2 %>%
  group_by(age) %>%
  summarize(mean_fracusing = mean(fracusing))
```

Similarly, group and summarize by education level:

```{r}
cuse2 %>%
  group_by(education) %>%
  summarize(mean_fracusing = mean(fracusing))
```

And group and summarize by desire for more children:

```{r}
cuse2 %>%
  group_by(wantsMore) %>%
  summarize(mean_fracusing = mean(fracusing))
```

Here the `%>%` is called a "pipe", and it sends the output of the previous function to the input of the next function. This could also have been done in one step:

```{r}
mutate(cuse, fracusing = using / (using + notUsing)) %>%
  group_by(age) %>%
  summarize(mean(fracusing))
```

Here the result is not stored anywhere; if you wanted to store it in a variable called `myanswer`, you could have started the above command with `myanswer <-` or `myanswer =`.

Also note that the line breaks are optional: you could put this all on one line, but breaking it up makes it more readable.

# Write on paper the model for expected probability of using birth control

Based on ```fit1```, write on paper the model for expected probability of using birth control? Don't forget the (inverse) logit function.

**Solution**

```{r}
fit1 <- glm(
  cbind(using, notUsing) ~ age + education + wantsMore,
  data = cuse,
  family = binomial("logit")
)
```

I'm going to store the coefficients of `fit1` in a new variable `b` just to save typing when writing out the formula, and round to two decimal places:

```{r}
bcoef <- round(coef(fit1), 2)
```

Here I'm going to take advantage of R Markdown's support for LaTeX formulae to write out the fitted regression model. I also use the \` r 2+2 \`     syntax (the result is `r 2+2`) for putting results of R code inline, instead of writing out numbers of the coefficients. I never wrote the number "four", but you'll have to look at the .Rmd to see how I did that! Doing this in reports allows you to update them as input data changes, without having to manually re-enter or copy numbers.

$P = \textit{logit}^{-1} \left( 
   `r bcoef["(Intercept)"]` + `r bcoef["age25-29"]` \times \textit{age25-29} + 
   `r bcoef["age30-39"]` \times \textit{age30-39} +
   `r bcoef["age40-49"]` \times \textit{age40-49} +
   `r bcoef["educationhigh"]` \times \textit{educationhigh} +
   `r bcoef["wantsMoreyes"]` \times \textit{wantsMoreyes} 
   \right)$

$\textit{logit}^{-1}(x) = \frac{1}{1+e^{-x}}$

# What is the expected probability...

Based on ```fit1```, what is the expected probability of an individual 25-29 years old, with high education, who wants more children, using birth control? Calculate it manually, and using `predict(fit1)`

**Solution**

This can be done using the `predict()` function, using a new `data.frame` giving the values that you want to predict for:
```{r}
invLogit <- function(x) 1 / (1 + exp(-x))
q4data <- data.frame(age = "25-29", education = "high", wantsMore = "yes")
invLogit(predict(fit1, newdata = q4data))
```

By default, `predict()` predicts on the original data, and taking a look, I noticed that the 7th value is the one we want to predict on:
```{r}
invLogit(predict(fit1)[7])
```

If we were really doing this manually, on paper, we would calculate the linear predictor ($\eta$), then evaluate the inverse logit function:
$$P = \textit{logit}^{-1}(\eta) = \frac{1}{1 + e^{-\eta}}$$

Note that `education` has reference level `"low"`, so for an individual with high education, the indicator variable `educationhigh` equals 1 and its coefficient must be included:

```{r}
eta <- bcoef["(Intercept)"] + bcoef["age25-29"] + bcoef["educationhigh"] + bcoef["wantsMoreyes"]
invLogit(eta)
```

Notice that this manual calculation matches our `predict()` result above (`r round(invLogit(eta), 3)`).

*(Tip: In `predict()`, specifying `type = "response"` directly returns predicted probabilities on the scale of the response variable, without needing a manual inverse-logit transformation: `predict(fit1, newdata = q4data, type = "response")`)*


# Predicted odds ratio


Based on ```fit1```: Relative to women under 25 who want to have children, what is the predicted odds ratio for a woman 40-49 years old who does _not_ want to have children will be taking birth control?

**Solution**

```{r}
exp(coef(fit1)["age40-49"] - coef(fit1)["wantsMoreyes"])
```

# Likelihood Ratio Test

Using a likelihood ratio test, is there evidence that a model with interactions improves on ```fit1``` (no interactions)?

**Solution**

```{r}
fit1.2way <- glm(
  cbind(using, notUsing) ~ (age + education + wantsMore)^2,
  data = cuse,
  family = binomial("logit")
)

fit1.int <- glm(
  cbind(using, notUsing) ~ age * education * wantsMore,
  data = cuse,
  family = binomial("logit")
)

anova(fit1, fit1.2way, fit1.int, test = "LRT")
```

The likelihood ratio test shows that adding all two-way interactions (`fit1.2way`) significantly improves model fit over the additive model (`fit1`, $p = 0.00027$). However, adding three-way interactions (`fit1.int`) yields no statistically significant improvement ($p = 0.486$) and costs 3 additional degrees of freedom (yielding a fully saturated model with 0 residual degrees of freedom). Therefore, the two-way interaction model is much more parsimonious.

# A model with interactions

Which, if any, variables have the strongest interactions? I use the "stargazer" package here which is extremely flexible for showing regression results from all kinds of models, capable of comparing multiple models or multiple dependent variables in the same table, and providing formatting customized to numerous journal styles.

**Solution**

In the two-way interaction model (`fit1.2way`), the strongest interaction coefficients are between age and wanting more children: `age30-39:wantsMoreyes` (`r round(coef(fit1.2way)["age30-39:wantsMoreyes"], 2)`) and `age40-49:wantsMoreyes` (`r round(coef(fit1.2way)["age40-49:wantsMoreyes"], 2)`). Both are statistically significant and negative ($p < 0.05$). This indicates that for women aged 30 and older, wanting more children exerts a substantially stronger deterrent effect against using birth control than would be predicted from age or wanting more children acting purely additively on the log-odds scale.

*(Note on the 3-way model: In the fully saturated model `fit1.int`, these interaction effects get partitioned across the three-way interaction terms, diluting the two-way coefficients and inflating standard errors because there are 16 parameters for 16 data rows.)*

The *stargazer* package is a good way to create nicely-formatted tables 
of regression results.

```{r, echo=FALSE, results="asis", message=FALSE}
library(stargazer)
stargazer(fit1, fit1.2way, fit1.int,
  type = "html",
  dep.var.caption = " ", # Otherwise the default is "Dependent Variable:"
  dep.var.labels = "Outcome: Using Birth Control",
  digits = 2, # show 2 digits
  column.labels = c("Additive", "Two-Way", "Full Three-Way"),
  ci = c(TRUE, TRUE, TRUE), # show confidence intervals
  single.row = TRUE, # put results on a single line
  report = "vc*s"
) # report the "v"ariable name, "c"oefficient with significance "asterisks", and "s"tandard error (actually replaced by confidence interval because of ci argument above)
```

# Contrasts between every pair of age groups

Create a model matrix for a fit on age only, with contrasts between *every pair* of age groups. Between which age groups is the contrast significant?

**Solution**

## semi-manually

First, I'll do it manually, using the `multcomp` package and following an example from `?glht`. I'm not going to bother anymore creating pretty tables with `stargazer` though... 

```{r}
fit <- glm(cbind(using, notUsing) ~ age, data = cuse, family = binomial("logit"))
coef(fit)
K <- rbind(
  "25-29 - <25" = c(0, 1, 0, 0),
  "30-39 - <25" = c(0, 0, 1, 0),
  "40-49 - <25" = c(0, 0, 0, 1),
  "30-39 - 25-29" = c(0, -1, 1, 0),
  "40-49 - 25-29" = c(0, -1, 0, 1),
  "40-49 - 30-39" = c(0, 0, -1, 1)
)
K
fit.all.cont <- multcomp::glht(fit, linfct = K)
summary(fit.all.cont)
```

## Using Tukey contrasts

There is an easier way if you realize that these are the "Tukey" contrasts and use the `contrMat` function from the `multcomp` package. These automatic contrasts
assume a no-intercept model, which would be hard to interpret outside the context of 
`multcomp::contrMat`, but can be used as an intermediate step to get here.

```{r}
fitnoint <- glm(cbind(using, notUsing) ~ age - 1, data = cuse, family = binomial("logit"))
K2 <- multcomp::contrMat(1:4, type = "Tukey")
K2
fit.all.cont2 <- multcomp::glht(fitnoint, linfct = K2)
summary(fit.all.cont2)
```

We didn't get the same informative coefficient names this time, but we can
rename the rows in `K2`, for example:

```{r}
rownames(K2) <- rownames(K)
K2
summary(multcomp::glht(fitnoint, linfct = K2))
```

> **Important Note on Contrasts and Intercepts:**
> Notice that `K2` must be tested using `fitnoint` (`multcomp::glht(fitnoint, linfct=K2)`), **not** `fit`. Because `contrMat()` constructs contrasts assuming each column corresponds to a separate group mean/log-odds ($\beta_1, \beta_2, \beta_3, \beta_4$ in a no-intercept model), applying `K2` to a model with an intercept (`fit`) mistakenly treats column 1 as the intercept ($\beta_0$) rather than the first group. The row `c(-1, 1, 0, 0)` would then compute $-\beta_0 + \beta_{25-29}$ (yielding $1.97$) instead of $\beta_{25-29} - \beta_{<25}$ ($0.46$). Using `fitnoint` ensures columns match the four group levels, correctly reproducing the exact estimates and p-values from our manual matrix `K` applied to `fit`.

### Conclusion: Between which age groups is the contrast significant?

Looking at the simultaneous tests with adjusted $p$-values:
- All pairwise differences in contraceptive use between age groups are statistically significant ($p < 0.05$), **except** between the **40–49** and **30–39** age groups (difference in log-odds = $0.376$, adjusted $p = 0.104$). Contraceptive use increases steadily with age from `<25` up to `30–39`, but plateaus between `30–39` and `40–49`.


## Other contrast schemes

There are plenty of canned contrast schemes provided by `multcomp::contrMat`. Note use of the `example()` function to run all examples from the `contrmat()` function. This is a great example of how there are **many** ways to analyze one particular experimental design, but that you need to know design matrices to utilize many of them. 

In these examples, the first line `n <- c(10,20,30,40)` just signifies four levels, `n = 1:4` would do exactly the same thing.

```{r, message=FALSE}
library(multcomp)
example("contrMat")
```

---

# Exercises

### Model Comparison & Parsimony via Likelihood Ratio Test

In the lab walkthrough, we compared the additive model (`fit1`) to the all-two-way model 
(`fit1.2way`) and the full three-way model (`fit1.int`). A researcher suggests that the 
only interaction of substantive interest is between age and desire for more children:

$$\text{cbind(using, notUsing)} \sim \text{age} \times \text{wantsMore} + \text{education}$$

1. Fit this model and name it `fit_agewants`.

```{r}

```

2. Conduct a Likelihood Ratio Test comparing the additive model (`fit1`) against 
`fit_agewants` using `anova(fit1, fit_agewants, test = "Chisq")`. Is the inclusion 
of the `age * wantsMore` interaction statistically justified?

```{r}

```

3. Compare `fit_agewants` against the model with all two-way interactions (`fit1.2way`) 
using `anova(fit_agewants, fit1.2way, test = "Chisq")`. Which model would you select as 
your final model? Why?

```{r}

```


