---
title: "Session 2 lab exercise: Exploration of contraceptive use data"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
---

**Learning Objectives**

1. Define "tidy" data
2. Load a dataset in R and perform basic exploratory data analysis
3. Create descriptive "Table 1" of a study sample using the `table1` package
4. Create a customized bar plot using the `ggplot2` package


**Activities**

1. Load the contraceptive use dataset into R
2. Create an "Epi Table 1" of sample characteristics
3. Create a barplot stratified by age and showing the relative proportions of 
participants using contraceptives among those who do and do not want more children.
4. Repeat the barplot showing percentages instead of counts

# Load the contraceptive use data

Load the data from https://github.com/waldronbios2/cunybios2/blob/main/session2/vignettes/cuse.dat. From this page:

> These data show the distribution of 1607 currently married and fecund women interviewed in the Fiji Fertility Survey, according to age, education, desire for more children and current use of contraception.

Unlike in the lecture, here I demonstrate reading the dataset using the *readr* package. It identifies some warnings because the header of this file has extra spaces at the end which appear like an extra column, making it appear like the dataset has 6 columns instead of 5. But the dataset reads with 5 columns as we want it to anyways, because I chose to skip the blank column 6 in the graphical interface (which added the code `X6 = col_skip()` below). *readr* is good for identifying problems like this in text data files.

```{r}
# traditional method for loading data:
# cuse <- read.table("cuse.dat", header=TRUE)
# Using readr package with "File - Import Dataset" and manually setting factor levels.
# Note, you don't have to write all this code by hand! It was produced by the File - Import Dataset helper.
library(readr)
library(readr)
cuse <-
  read_table(
    "cuse.dat",
    col_types = cols(
      age = col_factor(levels = c("<25", "25-29", "30-39", "40-49")),
      education = col_factor(levels = c("low", "high")),
      wantsMore = col_factor(levels = c("no", "yes")),
      X6 = col_skip()
    )
  )
```
Note, these warnings relate to the presence of an empty column in the dataset, 
which I dropped using `X6 = col_skip()`

```{r}
problems()
```

```{r}
summary(cuse)
```


# Create an epi "Table 1"

Here's a simple way to create the summary table that is required for your assignments,
and for any epidemiological analysis you do. The text format is fine, and convenient
for moving into another pubication or presentation document. See the [table1 vignette](https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html) 
for more examples. 

WARNING: the counts in this table are incorrect, because each row corresponds to 
a number of counts of women using and not using contraceptives. We do this just 
to demo the table1 package, then make a correct table below after reshaping
this dataset.

```{r, message = FALSE}
library(table1)
table1(~ ., data = cuse)
```
**Table 1: characteristics of the contraceptive use dataset. Note, these counts
correspond to number of rows in the dataset, _not_ numbers of participants**

# Definition of tidy data

*Excerpted from the [seminal paper by Hadley Wickam](https://vita.had.co.nz/papers/tidy-data.pdf)*

> Happy families are all alike; every unhappy family is unhappy in its own way - Leo Tolstoy

To summarize, in **tidy data**:
1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.

**Messy data** is any other arrangement of the data. (according to Hadley and the tidyverse!)

# Tidy the data

First let's do some work on the data to get it in shape for plotting. I do these 
one step at a time to show what's happening, but you could also do chain these
steps all together using the pipe operator (`%>%`).

First, group by age and whether the participant reports wanting more children, and 
sum the number wanting or not wanting more children in each of these groups. Also
rename "notUsing" to "not using" to make a nicer legend later.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
cusebyage0 <- group_by(cuse, age, education, wantsMore) %>%
    summarise(using = sum(using), "not using" = sum(notUsing))
print(cusebyage0)
```

Next, pivot this into a longer table by putting the "using" and "not using" columns
into a single column called "contraceptive". This pivot is necessary to make the data "tidy".


```{r}
cusebyage <- pivot_longer(cusebyage0,
      cols = using:"not using",
      values_to = "n",
      names_to = "contraceptive")
cusebyage
```

# A correct table 1

We can finally make a correct table1 with *one* more transformation, uncounting 
on the `n` column (using the tidyr package again). Also, we use `label` from
the `table1` package to make more informative and readable labels. For interest,
this time we stratify based on wanting more children, and add a caption. See
`?table` for more options.

```{r}
cuseverylong <- uncount(cusebyage, n) #from the tidyr package
label(cuseverylong$age) <- "Age (years)"
label(cuseverylong$education) <- "Education level"
label(cuseverylong$wantsMore) <- "Wants more children?"
label(cuseverylong$contraceptive) <- "Using contraceptives?"
table1(~ . | wantsMore, rowlabelhead = "Wants more children?", 
       caption = "Table 1: Participants in the contraceptive use study",
       data = cuseverylong)
```

Note, there are at least several other packages for making the epi "Table 1". Take 
a look to see if any of them appeal to you:
* [desctable](https://cran.r-project.org/web/packages/desctable/vignettes/desctable.html): powerful, provides a "download Excel" button in html output. 
* [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html): very powerful and flexible for formatting and features. Requires a little more 
code to get to your table 1.
* [tableone](https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html): simple and text-based, but works quite well and has options for exporting to Excel for 
use in a manuscript.

# Create a barplot

See the [Data Visualization with ggplot2 Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf) for help on enhancing this barplot.

Now, make a sort of fancy greyscale barplot using ggplot2. You can make a nice
plot without using nearly so many options, but I want to demonstrate the flexibility
of making a bar plot with ggplot2. 
```{r}
ggplot(cusebyage, aes(x = wantsMore, weight = n, fill = contraceptive)) +
  # create a stacked bar plot, where the values provided are counts/frequencies,
  # and use black outlines for the bars.
  geom_bar(position = "stack", stat = "count", color = "black") + 
  # use facet_grid to separate the plots by age group
  facet_grid(.~age, labeller = label_both) +
  labs(title = "Contraceptive usage counts",
       subtitle = "in Contraceptive Use dataset",
       caption = "Contraceptive use in study sample") + 
  xlab("Wants more children?") + 
  ylab("Number of Participants") +
  # there are lots of scale_fill_* options for automatic color schemes, but I 
  # just want to specify the colors manually here.
  scale_fill_manual(values=c("white", "grey")) +
  theme_bw()
```

**Figure 1: contraceptive use in the study sample.** Bar plot is organized by age 
group and stacked by self-report of whether participant wants more children. The 
fraction of women wanting more children decreases with age, becoming a minority in 
the 40-49 age group. One unexpected observation in this bar chart is that in the
<25 age group, those reporting wanting more children appear *more* likely to 
report using contraceptives. Is this the case? One way to make this more visually
clear would be to use percentages, instead of counts, on the vertical scale. Try
this, by changing `weight = n` to `weight = percent` to use the "percent" column
as heights instead of the "n" column. While you're at it, change the y label to
reflect this change.

*Note*: The caption option in *ggplot2* is suitable for smaller, embedded captions. But
for publication the caption usually needs to be in separated text.

# Exercises

## 1. Plot percentages instead of counts

Here's some code to calculate the percentages.

```{r}
cusebyage <- group_by(cusebyage, age, wantsMore) %>% 
    mutate(percent = n / sum(n) * 100)
cusebyage
```

Now, repeat the barplot, but showing percentages instead of counts.

## 2. Tidy a repeated-measures table

The table below is intentionally messy. It contains blood pressure and medication
measurements for two visits, but the visits and measurement types are encoded in
the column names.

```{r}
messy_bp <- tibble::tribble(
  ~participant, ~group,      ~week1_bp, ~week2_bp, ~week1_med, ~week2_med,
  "A01",        "control",          128,       126,        "no",        "no",
  "A02",        "control",          142,       138,       "yes",       "yes",
  "B01",        "treatment",        135,       129,        "no",        "no",
  "B02",        "treatment",        150,       144,       "yes",        "no"
)
messy_bp
```

1. Identify every non-tidy feature of `messy_bp`. Use the three tidy-data rules
  above in your explanation.
2. Tidy the data so that each row represents one participant at one visit. The
  result must have exactly these columns, in this order:
  `participant`, `group`, `week`, `blood_pressure`, and `medication`.
  Keep `blood_pressure` numeric and `medication` character. Because the two
  measurements have different types, you will need to handle that explicitly
  while pivoting or immediately afterward.
3. Name the result `tidy_bp`, then print it after sorting by `participant` and
  `week`.

*Hint*: You can reshape this in a single `pivot_longer()` call using the special
`".value"` sentinel in `names_to` (e.g., `names_to = c("week", ".value")` with
`names_sep = "_"`).

Your result is verifiable if it has eight rows, four unique participants, two
unique weeks, and no duplicate `participant`-`week` combinations. It should print
as follows (the order of rows and columns should match):

| participant | group | week | blood_pressure | medication |
|:--|:--|:--|--:|:--|
| A01 | control | week1 | 128 | no |
| A01 | control | week2 | 126 | no |
| A02 | control | week1 | 142 | yes |
| A02 | control | week2 | 138 | yes |
| B01 | treatment | week1 | 135 | no |
| B01 | treatment | week2 | 129 | no |
| B02 | treatment | week1 | 150 | yes |
| B02 | treatment | week2 | 144 | no |

You can verify the structure with:

```{r, eval=FALSE}
stopifnot(
  identical(names(tidy_bp),
                  c("participant", "group", "week", "blood_pressure", "medication")),
  nrow(tidy_bp) == 8,
  n_distinct(tidy_bp$participant) == 4,
  n_distinct(tidy_bp$week) == 2,
  nrow(distinct(tidy_bp, participant, week)) == nrow(tidy_bp)
)
```
