---
title: "Session 1 lab: Association between cholesterol and age"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
---

**Learning objectives**

1. Download and load a local tab-separated dataset (`cholesterol.tsv`) into R
2. Create exploratory scatterplots with regression lines using `ggplot2`
3. Fit and interpret multiple linear regression models with interaction terms
4. Evaluate regression assumptions using diagnostic residual plots
5. Compare nested models using Analysis of Variance (partial F-test)

# Downloading and Locating the Data File

To practice loading datasets from your local machine, download the `cholesterol.tsv` data file before proceeding:

* **Direct Download Link:** [cholesterol.tsv](https://raw.githubusercontent.com/waldronbios2/cunybios2/main/session1/vignettes/cholesterol.tsv)
* **GitHub Repository Location:** Inside the `session1/vignettes/` folder of the course repository.

### Steps to download and set up:

1. Right-click the link above and choose **"Save Link As..."** (or **"Download Linked File As..."**) to save `cholesterol.tsv` into your R working directory.
2. In R / RStudio, you can check your current working directory with:
   ```{r, eval=FALSE}
   getwd()
   ```
3. Alternatively, you can download the file directly into your working directory from R:
   ```{r, eval=FALSE}
   download.file(
     url = "https://raw.githubusercontent.com/waldronbios2/cunybios2/main/session1/vignettes/cholesterol.tsv",
     destfile = "cholesterol.tsv"
   )
   ```

# Load the dataset

You can practice loading this file using RStudio's graphical helper (**File → Import Dataset → From Text (readr)...**) or by writing the code directly using the `readr` package:

```{r, message=FALSE}
library(readr)
chol <- read_tsv("cholesterol.tsv", 
                 col_types = cols(
                   cholesterol = col_double(),
                   age = col_double(),
                   state = col_factor()
                 ))
summary(chol)
```

# Create a scatterplot of Cholesterol vs. Age

Explore the relationship between age and cholesterol levels, stratified by state:

```{r, message=FALSE}
library(ggplot2)
ggplot(chol, aes(x = age, y = cholesterol, shape = state, color = state)) + 
  geom_point(size = 4) +
  geom_smooth(method = lm, se = FALSE) +
  theme_bw() +
  labs(
    title = "Cholesterol vs. Age by State",
    x = "Age (years)",
    y = "Total Cholesterol (mg/dL)"
  )
```

# Fit a linear model with age, state, and interaction

Fit a multiple linear regression model with `cholesterol` as the continuous outcome, and `age`, `state`, and their interaction (`age * state`) as predictors:

```{r}
fit <- lm(cholesterol ~ age * state, data = chol)
summary(fit)
```

# Create an ANOVA table for this fit

```{r}
anova(fit)
```

# Diagnostic plots for the full model

Examine the standard regression diagnostic plots (Residuals vs Fitted, Normal Q-Q, Scale-Location, and Residuals vs Leverage):

```{r}
par(mfrow = c(2, 2))
plot(fit)
```

# Partial F-test for nested models

Compare a simpler model containing only `state` against a model containing both `state` and `age`:

```{r}
fit1 <- lm(cholesterol ~ state, data = chol)
fit2 <- lm(cholesterol ~ state + age, data = chol)
anova(fit1, fit2)
```

---

# Exercises

Now apply what you've learned to answer the following questions:

### Question 1: Testing for Interaction

Fit an additive model (`fit_additive`) without the interaction term:

```{r}

```


1. Use `anova(fit_additive, fit)` to conduct a partial F-test comparing the additive model to the interaction model.



2. Based on the p-value, is there statistically significant evidence that the slope (effect of age on cholesterol) differs between Iowa and Nebraska? Which model would you prefer?

### Question 2: Model Interpretation & Prediction

Using the additive model (`fit_additive`):

1. Interpret the regression coefficient for `age`: what is the estimated change in mean cholesterol for every 1-year increase in age, holding state constant?
2. What is the estimated difference in mean cholesterol between residents of Nebraska and Iowa of the same age?
3. Using the `predict()` function, predict the expected cholesterol level for:
   * A 50-year-old living in Iowa
   * A 50-year-old living in Nebraska

*Hint:*
```{r, eval=FALSE}
new_data <- data.frame(
  age = c(50, 50),
  state = factor(c("Iowa", "Nebraska"), levels = levels(chol$state))
)
predict(fit_additive, newdata = new_data)
```

### Question 3: Residual Diagnostics

Generate diagnostic plots for the additive model (`plot(fit_additive)`). Do the assumptions of linearity, equal variance (homoscedasticity), and normality of residuals appear reasonably satisfied?
