---
title: "Session 5 lab exercise: loglinear regression part 2"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
clean: false
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
  md_document:
    preserve_yaml: false
always_allow_html: true
---

# Learning objectives {-}

1. Fit Poisson, NB, and zero-inflated loglinear models
2. Perform nested deviance test for model selection
3. Make diagnostic plots of loglinear models
4. Visually assess multicollinearity among predictors

# Load the needle-sharing dataset

```{r loadandfilter, message=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
})
needledat <- readr::read_csv("needle_sharing.csv")
needledat2 <- needledat %>%
  dplyr::filter(sex %in% c("M", "F") &
                  ethn %in% c("White", "AA", "Hispanic") &
                  !is.na(homeless)) %>%
  mutate(
    homeless = recode_factor(homeless, "0" = "no", "1" = "yes"),
    hiv = recode_factor(
      hivstat,
      "0" = "negative",
      "1" = "positive",
      "2" = "positive"
    )
  )
```

# Compare the mean to the variance of the outcome variable. Calculate what fraction of the counts are zero.

```{r}
meanvarzeros <- filter(needledat2,!is.na(shared_syr)) %>%
  summarise(
    mean = mean(shared_syr),
    var = var(shared_syr),
    fraczero = sum(shared_syr == 0) / n()
  )
meanvarzeros
```

The mean number of needle sharing events per participant is `r signif(pull(meanvarzeros, mean), 3)`, the variance is `r signif(pull(meanvarzeros, var), 3)`, and the fraction of participants who never shared a needle is `r signif(pull(meanvarzeros, fraczero), 3)`. (Note how I put computed results in the text here rather than writing in numbers manually - they will change automatically if the analysis is changed!)

# Create a histogram of the outcome variable.

This was done in the lecture using base R, but let's do it here with ggplot2. Note the filtering of complete cases only is unnecessary because ggplot does it anyways, but this gets rid of a warning (try it without filtering). Specifying `binwidth = 1` gives each count its own bar; by default `geom_histogram` uses 30 bins spread across the range of the data, which here would group neighbouring counts together.

```{r}
library(ggplot2)
filter(needledat2, !is.na(shared_syr)) %>%
  ggplot(aes(x = shared_syr)) +
  geom_histogram(binwidth = 1)
```


# Fit Poisson and Negative Binomial models as in the lecture, with and without zero inflation.

## Poisson

```{r}
fit.pois <- glm(shared_syr ~ sex + ethn + homeless,
                data = needledat2,
                family = poisson(link = "log"))
```

## Negative Binomial

```{r}
library(MASS)
fit.negbin <- glm.nb(shared_syr ~ sex + ethn + homeless,
                data = needledat2)
```

## Zero-inflated Poisson

The formula has two parts separated by `|`: predictors of the count model go on the left, and predictors of the zero-inflation (logistic regression) model go on the right. `| 1` creates an intercept-only zero-inflation model; replace the `1` with any variables, using regular model formula syntax, to add them to the zero-inflation model. If you omit the `|` entirely, as in this first model, the same predictors are used for both parts.

```{r}
library(pscl)
fit.ZIpoisfull <- zeroinfl(shared_syr ~ sex + ethn + homeless,
                data = needledat2,
                dist = "poisson")
```

## Intercept-only ZI Poisson model

```{r}
fit.ZIpois <- zeroinfl(shared_syr ~ sex + ethn + homeless | 1,
                data = needledat2,
                dist = "poisson")
```

## Intercept-only ZI Negative Binomial model

```{r}
fit.ZInegbin <- zeroinfl(shared_syr ~ sex + ethn + homeless | 1,
                data = needledat2,
                dist = "negbin")
```

## Make a boxplot of needle sharing by homelessness (and other predictors)

Try replacing `homeless` with `ethn` or `sex`.

```{r}
filter(needledat2, !is.na(shared_syr)) %>%
  ggplot(aes(x = homeless, y = shared_syr)) +
  geom_boxplot(varwidth = TRUE)
```

# Assess whether there may be multicollinearity among the predictors

This uses `model.matrix` as a trick to produce a numeric matrix containing numeric and dummy variables, that can be used for calculating correlations and plotting heatmaps or dendrograms. First, assessing only three variables. Note, the first column of `mm` is the intercept, which must be removed. 

```{r}
mm <- model.matrix(~ sex + ethn + homeless, data = needledat2)
plot(hclust(as.dist(1 - cor(mm[, -1]))))
```

Now embed the dendrogram in a heatmap, which may be more or less informative, but worth trying. Here I use Euclidian distance for rows (patients) because some rows have zero variance and produce an error when attempting to calculate correlation, but use Pearson correlation for columns. The choice of distance metric makes this clustering look very different. See `?pheatmap` for other distance metric options. There are also other heatmap packages, but `?pheatmap` makes it easy to produce a basic heatmap.

```{r}
library(pheatmap)
pheatmap(mm[, -1],
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "euclidean")
```

Now repeat but assessing all the potential predictors. First remove `id` because it's just an identifier, the outcome and the variables derived from it (`shared_syr`, `shsyryn`, `logshsyr`, `shsyr`), and `hivstat` because it is recoded as `hiv`. These have to be removed with `select()` *before* calling `model.matrix()`: a formula like `~ . - shsyr` removes the term but still drops every row where `shsyr` is missing, which would leave only 22 of the 117 participants.

```{r}
predictors <- dplyr::select(needledat2, -id, -shared_syr, -shsyryn, -logshsyr, -shsyr, -hivstat)
mm <- model.matrix(~ ., data = predictors)[, -1] #[, -1] gets rid of intercept
nrow(mm) # participants with complete data on all predictors
plot(hclust(as.dist(1 - cor(mm))))
library(pheatmap)
pheatmap(mm,
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "correlation")
```


# Use chi-square nested deviance tests to assess which model seems to fit best.

I want to calculate the log-likelihood from each model. The simplest way is to call the `logLik` function one at a time:

```{r}
logLik(fit.pois)
logLik(fit.negbin)
logLik(fit.ZIpois)
logLik(fit.ZInegbin)
```

Just to demonstrate a fancier way that could be used on many models, I'll create a list of model objects:
```{r}
listoffits <-
  list(
    pois = fit.pois,
    negbin = fit.negbin,
    ZIpois = fit.ZIpois,
    ZInegbin = fit.ZInegbin
  )
```

Then demonstrate how the `lapply` (also related functions like `sapply`) can be used to implicitly loop over elements of a list, to do exactly the same thing:
```{r}
lapply(listoffits, logLik)
```

OK now to actually answer the question. Just to get an idea of how big a difference in $-2 \times log(likelihood)$ would be statistically significant on one difference of degrees of freedom:

```{r}
qchisq(0.95, df = 1)
```

Or two degrees of freedom:
```{r}
qchisq(0.95, df = 2)
```

All the differences in double log-likelihoods above are _much_ larger (in the hundreds) than these critical significance values, except for the difference between Negative Binomial and zero-inflated negative binomial models. So it doesn't look like zero inflation helped the Negative Binomial distribution model, but it helped the Poisson model, and the Negative Binomial model fits better than the Poisson model.

Strictly, the Poisson vs. Negative Binomial comparison needs a small correction. The Poisson model is the special case of the Negative Binomial where the dispersion parameter is zero, and since dispersion can't be negative, the null hypothesis sits at the boundary of the possible values. In that situation the likelihood ratio statistic doesn't follow a $\chi^2_1$ distribution: the correct p-value is half of the $\chi^2_1$ p-value, so the 5% critical value is the 90th percentile of $\chi^2_1$ rather than the 95th:

```{r}
qchisq(0.90, df = 1)
```

That makes no difference here, where the difference is in the hundreds, but it would matter for a borderline result.

Whether to fit a zero-inflated model at all is first a modeling question rather than a testing one. A zero-inflated model assumes there is a separate group of "structural zeros": here, people who would never share a syringe under any circumstances, as opposed to people who could have shared but happened not to in the past month. We have no particular reason to suspect such a group in this study, and the Negative Binomial model's overdispersion already accounts for the large number of zeros, which is consistent with zero inflation not improving the Negative Binomial model.

# Create residual deviance plots using the functions defined in the lecture.

These were the (base graphics) functions defined to create the first two panels of residuals plots for all of these types of models.
```{r}
plotpanel1 <- function(fit, ...) {
  plot(
    x = predict(fit),
    y = residuals(fit, type = "pearson"),
    xlab = "Predicted Values",
    ylab = "Pearson Residuals",
    ...
  )
  abline(h = 0, lty = 3)
  lines(lowess(x = predict(fit), y = resid(fit, type = "pearson")),
        col = "red")
}
plotpanel2 <- function(fit, ...) {
  resids <- scale(residuals(fit, type = "pearson"))
  qqnorm(resids, ylab = "Std Pearson resid.", ...)
  qqline(resids)
}
```

Let's make these plots. As a shortcut, remember that list of models? I'm going to use an explicit `for` loop this time instead of `lapply` so that I can use the vector names as plot titles.

Although we saw some evidence from the chi-square test that the Negative Binomial distribution fit better than the Poisson distribution (not surprising since these data are very over-dispersed), these diagnostic plots show the Negative Binomial distribution still does not fit well at all. I would take any interpretation of the coefficients of these models with plenty of skepticism. But this dataset is tricky and I'm not sure offhand of a good model to fit it.

Note, the line `par(mfrow=c(1, 2))` only works for base graphics (not ggplot2), and creates a 1 row by 2 column plot panel. 

```{r}
par(mfrow=c(1, 2))
for (i in seq_along(listoffits)){
  plotpanel1(listoffits[[i]], main = names(listoffits)[i])
  plotpanel2(listoffits[[i]], main = names(listoffits)[i])
}
```

# Plot predicted and observed counts

Here is a `data.frame` that we can use to make histograms, density plots, etc. Note `type = "response"`: for `glm` and `glm.nb` models, `predict()` returns predictions on the log (linear predictor) scale by default, whereas we want predicted counts that can be compared to the observed counts.
```{r}
preds <- data.frame(observed = fit.pois$y,
                    poisson = predict(fit.pois, type = "response"),
                    negbin = predict(fit.negbin, type = "response"),
                    ZIpois = predict(fit.ZIpois, type = "response"),
                    ZInegbin = predict(fit.ZInegbin, type = "response"))
```

Just to help with pivoting, let's pivot this into long-format:
```{r}
preds.long <- pivot_longer(preds, everything())
```

What did this do?
```{r}
summary(preds.long)
```

Boxplot
```{r}
ggplot(preds.long, aes(x = name, y = value)) + geom_boxplot()
```

Histogram with facet_wrap
```{r}
ggplot(preds.long, aes(x = value)) + 
  facet_wrap(~name) +
  geom_histogram(binwidth = 1)
```

We can see that none of the models come close to modeling the extreme observed counts of 30+. In reality, these might require a more complex mixture model: for example a mixture of zero-inflation plus two different count distributions, one with a much higher mean. This is beyond the scope of the course, but it is possible to fit more complex mixture models that could fit this dataset better.

# Exercises

Noting the caveat that the Negative Binomial model did not fit well, we will proceed to interpret it anyways. In reality, **bootstrap would be a much safer way to estimate confidence intervals and p-values**. If you wish, please feel free to substitute bootstrap in the following exercises, but it is not required.

## Exercise 1: Interpret Negative Binomial Coefficients as IRRs
Extract the coefficients from `fit.negbin`, exponentiate them to obtain Incidence Rate Ratios (IRRs), and compute their 95% confidence intervals (`confint()`). Interpret the IRR for `homeless`: how does needle sharing frequency compare between homeless and non-homeless participants, holding sex and ethnicity constant?

```{r}

```

## Exercise 2: Does Zero Inflation Depend on the Predictors?
The lab already fit `fit.ZIpoisfull`, a zero-inflated Poisson model where `sex`, `ethn`, and `homeless` predict *both* the count process and the zero-inflation process, but never used it. Compare `fit.ZIpoisfull` to the intercept-only zero-inflation model `fit.ZIpois` using a Likelihood Ratio Test. `anova()` doesn't work for `zeroinfl` models, so either use `lmtest::lrtest(fit.ZIpois, fit.ZIpoisfull)`, or calculate it manually as `2 * (logLik(fit.ZIpoisfull) - logLik(fit.ZIpois))` and compare it to `qchisq(0.95, df = ...)`, where `df` is the difference in the number of estimated parameters between the two models (see the `df` shown by `logLik()`). Do the predictors significantly improve the model of *who is a structural zero*, or is an intercept-only zero-inflation model adequate?

```{r}

```

## Exercise 3: Check for Overdispersion in the Poisson Model
Calculate the ratio of residual deviance to residual degrees of freedom for `fit.pois`. Given this value and the mean/variance comparison calculated earlier in the lab, explain why a Poisson model is a poor choice for this outcome, and why moving to a Negative Binomial model (as done in this lab) was the appropriate next step.

```{r}

```


